A model including an interaction of treatment with baseline risk was the optimal risk-based approach to predicting absolute treatment benefit

### 2021-09-06 submission/manuscript.rmd
add sample size restriction for adaptive method

Third, we considered a logistic regression model including treatment, the prognostic index, and their interaction. Absolute benefit is then estimated from 
$\hat{\tau}(\bm{x})=\text{expit}(\beta_0+\beta_{PI}PI) - \text{expit}(\beta_0+\beta_{t_x}+(\beta_{PI}+\beta_*)PI)$.
We will refer to this method as the *linear interaction* approach.

Finally, we used *restricted cubic splines* (RCS) to relax the linearity
assumption on the effect of the linear predictor [@Harrell1988]. We considered
splines with 3, 4 and 5 knots to compare models with different levels of
flexibility.

@@

```{r rmsebase, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated superpopulation of size 500,000. \\textit{Panel A} presents the results of the scenario with true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \\textit{Panel B} presents the results under moderate linear deviations from constant treatment effects; \\textit{Panel C} presents the results for strong linear deviations from constant relative treatment effects; \\textit{Panel D} presents the results for strong quadratic deviations from constant relative treatment effects.", fig.show="hold", out.width = '100%'}
knitr::include_graphics(here::here("figures/rmse_base.png"))
```

@@

```{r rmsesamplesize, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000. \\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 17,000; \\textit{Panel B} presents the results under moderate linear deviations from constant treatment effects; \\textit{Panel C} presents the results for strong linear deviations from constant relative treatment effects; \\textit{Panel D} presents the results for strong quadratic deviations constant relative treatment effects.", fig.show="hold", out.width = '100%'}
knitr::include_graphics(here::here("figures/rmse_sample_size.png"))
```

@@

```{r rmseauc, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000. \\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.85 and sample size of 4,250; \\textit{Panel B} presents the results under moderate linear deviations from constant treatment effects, while holding true prediction AUC and sample size constant; \\textit{Panel C} presents the results for strong linear deviations from constant relative treatment effects; \\textit{Panel D} presents the results for strong quadratic deviations from constant relative treatment effects.", fig.show="hold", out.width = '100%'}
knitr::include_graphics(here::here("figures/rmse_auc.png"))
```

@@

```{r discrimination, echo=FALSE, fig.cap="Discrimination for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000. \\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \\textit{Panel B} presents the results under moderate linear deviations from constant treatment effects; \\textit{Panel C} presents the results for strong linear deviations from constant relative treatment effects; \\textit{Panel D} presents the results for strong quadratic deviations from constant relative treatment effects.", fig.show="hold", out.width = '100%'}
knitr::include_graphics(here::here("figures/discrimination_base.png"))
```

@@

```{r calibration, echo=FALSE, fig.cap="Calibration for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000. \\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \\textit{Panel B} presents the results under moderate linear deviations from constant treatment effectst; \\textit{Panel C} presents the results for strong linear deviations from of constant relative treatment effects; \\textit{Panel D} presents the results for strong quadratic deviations from constant relative treatment effects.", fig.show="hold", out.width = '100%'}
knitr::include_graphics(here::here("figures/calibration_base.png"))
```

@@

Including restricted cubic splines with 4 or 5 knots proved to be too flexible,
as indicated by higher RMSE, increased variability of discrimination for benefit
and worse calibration of benefit predictions. Even with larger sample sizes and
strong quadratic deviations from the base case scenario of constant relative
treatment effects, these more flexible restricted cubic splines did not
outperform the simpler RCS with 3 knots. These more flexible approaches may only
be helpful if we expect more extreme patterns of heterogeneous treatment
effects.

@@

Increasing the discriminative ability of the risk model – by increasing the
predictor coefficients of the true risk model – reduced RMSE for all methods.
This increase in discriminative ability translates in higher variability of
predicted risks, which, in turn, allows the considered methods to better capture
absolute treatment benefits. As a consequence, the increase in discriminative
ability of the risk model also led to higher values of c-for-benefit. Even
though risk model performance is very important for the ability of risk-based
methods to predict treatment benefit, prediction model development was outside
the scope of this work and has already been studied extensively
[@vanKlaveren2019; @Burke2014; @Abadie2018].

>>>>

Third, we considered a logistic regression model including treatment, the
prognostic index, and their linear interaction. Absolute benefit is then
estimated from
$\hat{\tau}(\bm{x})=\text{expit}(\beta_0+\beta_{PI}PI) - \text{expit}(\beta_0+\beta_{t_x}+(\beta_{PI}+\beta_*)PI)$.
We will refer to this method as the *linear interaction* approach.

Fourth, we used *restricted cubic splines* (RCS) to relax the linearity
assumption on the effect of the linear predictor [@Harrell1988]. We considered
splines with 3, 4 and 5 knots to compare models with different levels of
flexibility.

Finally, we considered an adaptive approach using Akaike's Information Criterion
(AIC) for model selection. The candidate models were: a constant treatment
effect model, a model with a linear interaction with treatment and RCS models
with 3, 4 and 5 knots.

@@

```{r rmsebase, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated superpopulation of size 500,000. The scenario with true constant relative treatment effect had a true prediction AUC of 0.75 and sample size of 4250. Results are presented under moderate linear, strong linear, and strong quadratic deviations from constant relative treatment effects.", fig.show="hold", out.width = '100%'}
knitr::include_graphics(here::here("figures/rmse_base.png"))
```

@@

```{r rmsesamplesize, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000. Sample size 17,000 rather than 4250 in Figure \\ref{fig:rmsebase}", fig.show="hold", out.width = '100%'}
knitr::include_graphics(here::here("figures/rmse_sample_size.png"))
```

@@

```{r rmseauc, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000. True prediction AUC of 0.85 and sample size of 4,250.", fig.show="hold", out.width = '100%'}
knitr::include_graphics(here::here("figures/rmse_auc.png"))
```

@@

```{r discrimination, echo=FALSE, fig.cap="Discrimination for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000. True prediction AUC of 0.75 and sample size of 4,250.", fig.show="hold", out.width = '100%'}
knitr::include_graphics(here::here("figures/discrimination_base.png"))
```

@@

```{r calibration, echo=FALSE, fig.cap="Calibration for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000. True prediction AUC of 0.75 and sample size of 4,250.", fig.show="hold", out.width = '100%'}
knitr::include_graphics(here::here("figures/calibration_base.png"))
```

@@

Models including restricted cubic splines with 4 or 5 knots proved to be too
flexible, as indicated by higher RMSE, increased variability of discrimination
for benefit and worse calibration of benefit predictions. Even with larger
sample sizes and strong quadratic deviations from the base case scenario of
constant relative treatment effects, these more flexible restricted cubic
splines did not outperform the simpler RCS with 3 knots. These approaches may
only be helpful if we expect more extreme patterns of heterogeneous treatment
effects compared to the quadratic deviations considered here.

@@

Increasing the discriminative ability of the risk model--by increasing the
predictor coefficients of the true risk model--reduced RMSE for all methods.
This increase in discriminative ability translates in higher variability of
predicted risks, which, in turn, allows the considered methods to better capture
absolute treatment benefits. As a consequence, the increase in discriminative
ability of the risk model also led to higher values of c-for-benefit. Even
though risk model performance is very important for the ability of risk-based
methods to predict treatment benefit, prediction model development was outside
the scope of this work and has already been studied extensively
[@vanKlaveren2019; @Burke2014; @Abadie2018].

The adaptive approach had adequate performance, following closely the
performance of the "true" model in most scenarios. However, with smaller sample
sizes it tended to "miss" the treatment-risk interactions and selected simpler
models. This resulted in increased RMSE variability in these scenarios,
especially in the case of true strong linear deviations from the base case
scenario. This tendency was mitigated when the discriminative ability of the
prediction model was higher. Therefore, in the case of smaller sample sizes
and/or poorly discriminating prediction models, the simpler linear interaction
model is a safer choice for predicting absolute benefits.

### 2021-09-06 submission/manuscript.tex
add sample size restriction for adaptive method

Third, we considered a logistic regression model including treatment,
the prognostic index, and their interaction. Absolute benefit is then
estimated from
\(\hat{\tau}(\bm{x})=\text{expit}(\beta_0+\beta_{PI}PI) - \text{expit}(\beta_0+\beta_{t_x}+(\beta_{PI}+\beta_*)PI)\).
We will refer to this method as the \emph{linear interaction} approach.

Finally, we used \emph{restricted cubic splines} (RCS) to relax the
linearity assumption on the effect of the linear predictor {[}7{]}. We
considered splines with 3, 4 and 5 knots to compare models with
different levels of flexibility.

@@

\begin{figure}
\includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/rmse_base} \caption{RMSE of the considered methods across 500 replications calculated in a simulated superpopulation of size 500,000. \textit{Panel A} presents the results of the scenario with true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \textit{Panel B} presents the results under moderate linear deviations from constant treatment effects; \textit{Panel C} presents the results for strong linear deviations from constant relative treatment effects; \textit{Panel D} presents the results for strong quadratic deviations from constant relative treatment effects.}\label{fig:rmsebase}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/rmse_sample_size} \caption{RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000. \textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 17,000; \textit{Panel B} presents the results under moderate linear deviations from constant treatment effects; \textit{Panel C} presents the results for strong linear deviations from constant relative treatment effects; \textit{Panel D} presents the results for strong quadratic deviations constant relative treatment effects.}\label{fig:rmsesamplesize}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/rmse_auc} \caption{RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000. \textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.85 and sample size of 4,250; \textit{Panel B} presents the results under moderate linear deviations from constant treatment effects, while holding true prediction AUC and sample size constant; \textit{Panel C} presents the results for strong linear deviations from constant relative treatment effects; \textit{Panel D} presents the results for strong quadratic deviations from constant relative treatment effects.}\label{fig:rmseauc}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/discrimination_base} \caption{Discrimination for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000. \textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \textit{Panel B} presents the results under moderate linear deviations from constant treatment effects; \textit{Panel C} presents the results for strong linear deviations from constant relative treatment effects; \textit{Panel D} presents the results for strong quadratic deviations from constant relative treatment effects.}\label{fig:discrimination}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/calibration_base} \caption{Calibration for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000. \textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \textit{Panel B} presents the results under moderate linear deviations from constant treatment effectst; \textit{Panel C} presents the results for strong linear deviations from of constant relative treatment effects; \textit{Panel D} presents the results for strong quadratic deviations from constant relative treatment effects.}\label{fig:calibration}
\end{figure}

@@

Increasing the discriminative ability of the risk model -- by increasing
the predictor coefficients of the true risk model -- reduced RMSE for
all methods. This increase in discriminative ability translates in
higher variability of predicted risks, which, in turn, allows the
considered methods to better capture absolute treatment benefits. As a
consequence, the increase in discriminative ability of the risk model
also led to higher values of c-for-benefit. Even though risk model
performance is very important for the ability of risk-based methods to
predict treatment benefit, prediction model development was outside the
scope of this work and has already been studied extensively
{[}5,12,13{]}.

>>>>

Third, we considered a logistic regression model including treatment,
the prognostic index, and their linear interaction. Absolute benefit is
then estimated from
\(\hat{\tau}(\bm{x})=\text{expit}(\beta_0+\beta_{PI}PI) - \text{expit}(\beta_0+\beta_{t_x}+(\beta_{PI}+\beta_*)PI)\).
We will refer to this method as the \emph{linear interaction} approach.

Fourth, we used \emph{restricted cubic splines} (RCS) to relax the
linearity assumption on the effect of the linear predictor {[}7{]}. We
considered splines with 3, 4 and 5 knots to compare models with
different levels of flexibility.

Finally, we considered an adaptive approach using Akaike's Information
Criterion (AIC) for model selection. The candidate models were: a
constant treatment effect model, a model with a linear interaction with
treatment and RCS models with 3, 4 and 5 knots.

@@

\begin{figure}
\includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/rmse_base} \caption{RMSE of the considered methods across 500 replications calculated in a simulated superpopulation of size 500,000. The scenario with true constant relative treatment effect had a true prediction AUC of 0.75 and sample size of 4250. Results are presented under moderate linear, strong linear, and strong quadratic deviations from constant relative treatment effects.}\label{fig:rmsebase}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/rmse_sample_size} \caption{RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000. Sample size 17,000 rather than 4250 in Figure \ref{fig:rmsebase}}\label{fig:rmsesamplesize}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/rmse_auc} \caption{RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000. True prediction AUC of 0.85 and sample size of 4,250.}\label{fig:rmseauc}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/discrimination_base} \caption{Discrimination for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000. True prediction AUC of 0.75 and sample size of 4,250.}\label{fig:discrimination}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/calibration_base} \caption{Calibration for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000. True prediction AUC of 0.75 and sample size of 4,250.}\label{fig:calibration}
\end{figure}

@@

Increasing the discriminative ability of the risk model--by increasing
the predictor coefficients of the true risk model--reduced RMSE for all
methods. This increase in discriminative ability translates in higher
variability of predicted risks, which, in turn, allows the considered
methods to better capture absolute treatment benefits. As a consequence,
the increase in discriminative ability of the risk model also led to
higher values of c-for-benefit. Even though risk model performance is
very important for the ability of risk-based methods to predict
treatment benefit, prediction model development was outside the scope of
this work and has already been studied extensively {[}5,12,13{]}.

The adaptive approach had adequate performance, following closely the
performance of the ``true'' model in most scenarios. However, with
smaller sample sizes it tended to ``miss'' the treatment-risk
interactions and selected simpler models. This resulted in increased
RMSE variability in these scenarios, especially in the case of true
strong linear deviations from the base case scenario. This tendency was
mitigated when the discriminative ability of the prediction model was
higher. Therefore, in the case of smaller sample sizes and/or poorly
discriminating prediction models, the simpler linear interaction model
is a safer choice for predicting absolute benefits.
